Index measures for oak decline severity using phenotypic descriptors

### 2019-12-14 manuscript/manuscript.Rmd
fixed feature importance plots

Oak declines are complex syndromes that are facilitated by a variety of both abiotic and biotic factors.
Abiotic factors can include both predisposing and inciting factors such as drought and are often
Biotic agents can include insect infestation such as the larvae of bark boring bettle *Agrilus biguttatus* associated

@@

It is becoming increasingly important that oak tree health within woodlands is accurately and sensitively monitoried to enable the detection of decline as well as it's spread and severity over time to enable accurate forecasting and inform forestry management practices (ref).
Monitoring methods need to take advantage of modern data analysis techniques whilst still retaining practical simplicity in the field.

@@

Phenotypic data were collected from a total of 174 trees from nine sites from across the south, east and west of England between June and September in 2016 or 2017.
The occurance of AOD has previously been reported and studied at six of the sites [@brown2016; @brown2017].

@@

**Crown production efficiency** - Estimate of crown production effieceny as a ratio between the composite crown surface area and the composite crown volumne [@zarnoch2004].

@@

The phenotypic descriptors were analysed using 100 repetitions of unsupervised random forest using the randomForest R package v`r packageVersion('randomForest') %>% as.character()` [@liaw2002].
The proximity matrices from each repetition were averaged to a single proximity matrix ($P$).
This was then transformed to a dissimilarity matrix ($1 - P$) and scaled to two dimensions using  multidimensional scaling (MDS).
The first MDS dimension was min-max scaled to give a scale between 0 and 1 with more severely declining trees having a score closer to 1.
This dimension was named the PDI.
The second MDS dimension ($D$) was first inverted ($1 - D$) to ensure the correct dimension orientation relative to decline type.
The dimension was then min-max scaled and then transformed ($2D - 1$) to give a scale between -1 and 1 with COD trees having a score closer to -1 and AOD having a score closer to 1. 
This dimension was named the DAI.
With different data, it is likely that the MDS dimensions may require inversion prior to min-max scaling to ensure their correct orientation.

@@

- simulated descriptor response surfaces

@@

Local Interpretable Model-Agnostic Explanations (LIME) To understand the descriptors contributing to case specific examples of each of the indexes,  were computed for a range of specific cases [@ribeiro2016]. 
The lime R package v`r packageVersion('lime') %>% as.character()` was used with the default parameters [@pedersen2019].
A low, a moderate and a severe decline cases was used for the PDI. For each case, a tree with a score closest to 0, 0.5 and 1 respectively was selected.
For the DAI, a decline neautral case was selected with a score closes to 0.
Then for each of the AOD and COD syndrome types a moderatefor the DAI .

@@

Trees with a score around 0 are neither distinctively AOD or COD sharing characteristics of both conditions in trees with PDI greater than 0.3.
A significant difference was found in the PDI between non-symptomatic (control) and symptomatic (AOD and COD) decline statuses using an independant two sample t-test (`r glue("t({PDI_status_ttest$parameter}) = {signif(PDI_status_ttest$statistic,digits = 3) %>% abs()}, p < 0.001")`) with group means of `r signif(PDI_status_ttest$estimate1,digits = 3)` and `r signif(PDI_status_ttest$estimate2,digits = 3)` for non-symptomatic and symptomatic statuses respectively.
Significance was also found in the DAI between the AOD and COD groups (`r glue("t({DAI_groups_ttest$parameter}) = {signif(DAI_groups_ttest$statistic,digits = 3) %>% abs()}, p < 0.001")`) with group means of `r signif(DAI_groups_ttest$estimate1,digits = 3)` and `r signif(DAI_groups_ttest$estimate2,digits = 3)` for the AOD and COD groups respectively.
A scatter plot of PDI against DAI is shown in Figure # for all 174 trees.
It can clearly be seen that the

@@

- Accuracies of retrained models
- Lime results
- Top ranked descriptor index response surfaces

@@

Crown condition descriptors such as crown volume and crown condition contributed highly to the PDI. 
The phenotypic measures central to this were percentage missing crown and percentage crowntransparency from which crown condition and crown volume are calculated. 
It can be seen in Figure 1 that as the percentage missing crown and percentage crown transparency increase and therefore crown condition and estimated crown volume decrease, the PDI increases. 
A much greater range of PDI values can be obtained from trees with relatively greater stature (e.g. greater crown radius, total height, lower crown height and diameter at breast height; Figures 1a & b), with relatively small stature trees having PDIs of 0.5 or greater even with 100% crown condition.

@@

- Accessability of collecting descriptors
- Suitability of collected phenotypic descriptors for describing decline status
- Concession of model interpretability to predictive capability by using a machine learning approach
- Trends of key phenotypic descriptors withing the PDI and DAI measures
- Sensitivity provided for integration with large scale omics analyses as well as long term monitoring for forest condition
- Index thresholds for category definitions (healthy, declining, severe decline etc.)
- Semi-supervised can be used if unsupervised model is unable to pull out decline severity scale
- Scalability of method for wider application
- Thresholds for decline groupings within index ranges
- Validation of decline scores - chicken and paradox
- Refined definition of remission
- use of unsupervised over supervised analysis
- Scale inversion likely need with new data

@@

- Discussion of applications to forestry monitoring and management
- Sensitivity for integration with molecular data compared to categorical assignments

>>>>

Oak declines are complex syndromes that are facilitated by a variety of both abiotic and biotic factors.
Abiotic factors can include both predisposing and inciting factors such as drought and are often
Biotic agents can include insect infestation such as the larvae of bark boring beetle *Agrilus biguttatus* associated

@@

It is becoming increasingly important that oak tree health within woodlands is accurately and sensitively monitored to enable the detection of decline as well as it's spread and severity over time to enable accurate forecasting and inform forestry management practices (ref).
Monitoring methods need to take advantage of modern data analysis techniques whilst still retaining practical simplicity in the field.

@@

Phenotypic data were collected from a total of 174 trees from nine sites from across the south, east and west of England between June and September in 2016 or 2017.
The occurrence of AOD has previously been reported and studied at six of the sites [@brown2016; @brown2017].

@@

**Crown production efficiency** - Estimate of crown production efficiency as a ratio between the composite crown surface area and the composite crown volume [@zarnoch2004].

@@

The phenotypic descriptors were analysed using 100 repetitions of unsupervised random forest using the randomForest R package v`r packageVersion('randomForest') %>% as.character()` [@liaw2002].
The proximity matrices from each repetition were averaged to a single proximity matrix ($P$).
This was then transformed to a dissimilarity matrix ($1 - P$) and scaled to two dimensions using  multidimensional scaling (MDS).
The first MDS dimension ($D$) was first inverted ($1 - D$) to ensure the correct dimension orientation relative to decline severity.
It was then min-max scaled to give a scale between 0 and 1 with more severely declining trees having a score closer to 1.
This dimension was named the PDI.
The second MDS dimension was similarly inverted to ensure the correct dimension orientation relative to decline type.
The dimension was then min-max scaled and then transformed ($2D - 1$) to give a scale between -1 and 1 with COD trees having a score closer to -1 and AOD having a score closer to 1. 
This dimension was named the DAI.
With different data, it is likely that the MDS dimensions may require inversion prior to min-max scaling to ensure their correct orientation.

@@

- descriptor importance for each index
- simulated descriptor response surfaces

@@

Overall descriptor importance contributions to each of the indexes was retrieved from these models using gini impurity (ref).
The assoication of key raw descriptors identified from the overall importance were investigated by simulating

Local Interpretable Model-Agnostic Explanations (LIME) were computed to identify descriptors contributing locally to case specific examples of each of the indexes [@ribeiro2016]. 
The lime R package v`r packageVersion('lime') %>% as.character()` was used with the default parameters [@pedersen2019].
A low, a moderate and a severe decline cases was used for the PDI. 
For each case, a tree with a score closest to 0, 0.5 and 1 respectively was selected.
For the DAI, a decline neutral case was selected with a score closes to 0.
Then for each of the AOD and COD syndromes, moderate and severe cases were selected with scores closes to 0.5 and 1 for AOD and -0.5 and -1 for COD.

@@

Trees with a score around 0 are neither distinctively AOD or COD sharing characteristics of both conditions in trees with PDI greater than 0.3.
A significant difference was found in the PDI between non-symptomatic (control) and symptomatic (AOD and COD) decline statuses using an independent two sample t-test (`r glue("t({PDI_status_ttest$parameter}) = {signif(PDI_status_ttest$statistic,digits = 3) %>% abs()}, p < 0.001")`) with group means of `r signif(PDI_status_ttest$estimate1,digits = 3)` and `r signif(PDI_status_ttest$estimate2,digits = 3)` for non-symptomatic and symptomatic statuses respectively.
Significance was also found in the DAI between the AOD and COD groups (`r glue("t({DAI_groups_ttest$parameter}) = {signif(DAI_groups_ttest$statistic,digits = 3) %>% abs()}, p < 0.001")`) with group means of `r signif(DAI_groups_ttest$estimate1,digits = 3)` and `r signif(DAI_groups_ttest$estimate2,digits = 3)` for the AOD and COD groups respectively.
A scatter plot of PDI against DAI is shown in Figure # for all 174 trees.
It can clearly be seen that the

@@

- Trends of key phenotypic descriptors withing the PDI and DAI measures
- Agrilus exit hole density had no contribution
- Lime results
- Top ranked descriptor index response surfaces

@@

Crown condition descriptors such as crown volume and crown condition contributed highly to the PDI. 
The phenotypic measures central to this were percentage missing crown and percentage crown transparency from which crown condition and crown volume are calculated. 
It can be seen in Figure 1 that as the percentage missing crown and percentage crown transparency increase and therefore crown condition and estimated crown volume decrease, the PDI increases. 
A much greater range of PDI values can be obtained from trees with relatively greater stature (e.g. greater crown radius, total height, lower crown height and diameter at breast height; Figures 1a & b), with relatively small stature trees having PDIs of 0.5 or greater even with 100% crown condition.

@@

- reasons for selecting machine learing approach
- use of unsupervised over supervised analysis
- Semi-supervised can be used if unsupervised model is unable to pull out decline severity scale
- For given context, descriptors not contributing to the indexes could be removed

- Accessability of collecting descriptors
- Suitability of collected phenotypic descriptors for describing decline status
- Concession of model interpretability to predictive capability by using a machine learning approach
- Index thresholds for category definitions (healthy, declining, severe decline etc.)
- Scalability of method for wider application
- Thresholds for decline groupings within index ranges
- Validation of decline scores - chicken and paradox

@@

- Discussion of applications to forestry monitoring and management
- Sensitivity for integration with molecular data compared to categorical assignments
- Sensitivity provided for integration with large scale omics analyses as well as long term monitoring for forest condition
- Refined definition of remission - repeat measurement
